---
title: "geotrellis: spatial data analysis using geotrellis"
date: "`r format(Sys.time(), '%d %B %Y')`"
output:
  rmarkdown::html_document:
    toc: false
    fig_caption: yes
fontsize: 11pt
vignette: >
  %\VignetteIndexEntry{geotrellis}
  %\VignetteEngine{knitr::rmarkdown}
  \usepackage[utf8]{inputenc}
---


```{r, include=FALSE}
devtools::load_all()
```

The _geotrellis_ _R_ package was an experiment to see if _R_ could leverage the [_Scala_](https://www.scala-lang.org) [data processing engine _geotrellis_](http://geotrellis.io) to process spatial data faster than the [_raster_](https://cran.r-project.org/package=raster) _R_ package. This vignette outlines the functions implmented in the geotrellis _R_ package, and provides some benchmark results against the _raster_ _R_ package. **This _R_ package is experimental and not recomended for real analysis. Current benchmarks indicate that little to no benefits will be gained using this package.**

Spatial data is stored in a `gt_RasterLayer` object. This _R6_ class contains all low-level the definitions for reading, writing, processing, and analyzing data. However, users should not interact with this class' methods directly. Instead, users should use the _S4_ methods provided. The usage of these _S4_ methods is outlined below. 

# Reading/write data

Spatial data is read using the `gt_raster` method. This method can load data using a `character` file path for a GeoTiff file. Additionally, users can coerce a `raster::RasterLayer-class` object. Note that the package only support single-band layer raster data. 

```{r}
# load in data using file path
g <- gt_raster(system.file('extdata/test.tif', package='geotrellis'))

# load in data using a RasterLayer object
g <- gt_raster(raster(system.file('extdata/test.tif', package='geotrellis')))
````

Once data has been loaded into _R_, users can access the metadata associated with the data. The methods used to access the metadata follow the conventions used in the _raster R_ package.

```{r}
print(g)  # print summary of object
crs(g)    # coordinate reference system
extent(g) # spatial extent of data
res(g)    # resolution
ncell(g)  # number of data cells
ncol(g)   # number of columns
nrow(g)   # number of rows
```

Users can also convert the data into native _R_ objects and `raster::RasterLayer-class` objects.

```{r}
values(g)    # convert data as a numeric vector
as.matrix(g) # convert data to a matrix
as.raster(g) # convert data to a RasterLayer
```

Finally, data can be saved to disk using the `gt_writeRaster` function.

```{r}
# save data to disk in a GeoTiff format
gt_writeRaster(g, tempfile(fileext='.tif')) 
```

# Geoprocessing

The _geotrellis R_ package contains several geoprocessing methods.

* To spatially align data in the same coordinate reference system, the `gt_resample` can be used:

```{r}
# create new layer with different spatial resolution
l <- gt_raster(raster::disaggregate(as.raster(g), 2, method=''))

# resample data to new resolution
gt_resample(g, l, method='ngb')
```

* To spatially project data to a different coordinate reference system, the `gt_projectRaster` method can be used. In addition to an input `gt_RasterLayer` object, this method accepts either: (1) a `sp::CRS` coordinate refence system object and a `numeric` resolution or (2) a `gt_RasterLayer` with spatial properties that are used to reproject the data.

```{r}
gt_projectRaster(g, sp::CRS('+init=epsg:3395'), res=50000, method='ngb')
```

* To extract contiguous chunk from a `gt_RasterLayer` object, the `gt_crop` method can be used. This method accepts either an `raster::Extent-class` object or a `gt_RasterLayer` object.

```{r}
gt_crop(g, extent(raster::extent(c(0, 3, 2, 6))))
```

* To set cells in in a `gt_RasterLayer` object to another value, the `gt_mask` method can be used. This defaults to updating cells in the argument to `x` to be `NA` where they are `NA` in the mask layer. The `maskvalue` and `updatevalue` arguments can used to modify this behavior though.

```{r}
# create mask layer
m <- as.raster(g)
m <- raster::setValues(m, sample(c(1,NA), ncell(m), replace=TRUE))
m <- gt_raster(m)

# run mask analysis
gt_mask(g, m)
```

# Statistics 

This package offers two statistical methods for analyzing data in a `gt_RasterLayer` object. Note that there is currently a bug in the geotrellis code that causes medians to be calculated incorrectly.  Additionally, note that unlike the functions in the _raster R_ package, these functions calculate the population standard deviation, not the sample standard deviation.

The `gt_cellStats` method can be used to calculate statistics using all the cells in the raster data. 

```{r}
gt_cellStats(g)
```

The the `gt_zonal` method can be used to calculate statistics in zones, as specified in a second raster.

```{r}
# make layer with zones
z <- as.raster(g)
z <- raster::setValues(z, sample(1:2, ncell(z), replace=TRUE))
z <- gt_raster(z)

# run zonal statistics
gt_zonal(g, z)
```

# Benchmarks

Finally, the _geotrellis R_ package provides a `benchmark` function to automate the benchmarking of all these functions and compare them to their counterparts in the _raster_ package. 

```{r}
# run benchmark
b <- benchmark(ncell=c(1e+1, 1e+2, 1e+3, 1e+4, 1e+5), times=10L)
# print benchmark
print(b)
# plot benchmark
plot(b)
```

# Acknowledgments

I would like to thank to the _geotrellis_ developers for their assistance.
